---
title: "Sound generation with soundgen"
author: "Andrey Anikin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sound generation with soundgen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Intro
## Purpose
The functions for sound synthesis in soundgen are intended for synthesis of non-linguistic vocalizations of both humans and other animals: sighs, moans, screams, etc. They can also be used for the creation of non-biological sounds that require precise control over spectral and temporal modulations, such as special sound effects in computer games or acoustic stimuli for scientific experiments. Soundgen is NOT meant to be used for text-to-speech conversion. It can be adapted for this purpose, but existing specialized tools will probably serve better.

Soundgen uses a parametric algorithm, which means that sounds are synthesized de novo using control parameters, as opposed to concatenating or modifying existing audio recordings. Under the hood, the current version of soundgen generates and filters two sources of excitation: sine waves for harmonics of tonal sounds and white noise for non-harmonic sounds. 

The rest of this vignette will unpack this last statement and demonstrate how soundgen can be used in practice. To simplify setting the control parameters and visualizing the output, soundgen library includes an interactive Shiny app. To start the app, type `soundgen_app()` from R or try it online at <a href="http://cogsci.se/soundgen.html">cogsci.se/soundgen.html</a>. To generate sounds from the console, use the function `soundgen()`. Each section of the vignette describes a particular aspect of sound generation, both describing the relevant arguments of `soundgen()` and explaining how they can be specified in the app.

## Before you proceed: consider the alternatives
Soundgen is a fairly complicated tool, so expect a learning curve. If your task is acoustically straightforward, you may want to look at some alternatives. In R, there are at least three other packages that offer sound synthesis: `tuneR`, `seewave`, and `phonTools`. Both `seewave` and `tuneR` implement straightforward ways to synthesize pulses and square, triangular, or sine waves as well as noise with adjustable (linear) spectral slope. You can also create multiple harmonics with both amplitude and frequency modulation using `seewave` functions `synth()` and `synth2()`. There is even a function available for adding formants and thus creating different vowels: `vowelsynth()` from the `phonTools` package. If this is ample for your needs, try these packages first. 

So why bother with soundgen at all? For the same reasons it was develloped in the first place. First, it takes customization and flexibility of sound synthesis much further. You will appreciate this flexibility if your aim is to produce convincing biological sounds. And second, it's a much higher-level tool with dedicated separate subroutines for things like controlling the rolloff (relative energy of different harmonics), adding moving formants and antiformants, mixing harmonic and noise components, controlling voice changes over multiple syllables, adding stochasticity to imitate unpredictable voice changes common in biological sound production, and more. In other words, soundgen offers powerful control over low-level acousic characteristics of synthesized sounds with the benefit of also offering transparent, meaningful high-level parameters intended for the creation of whole bouts of vocalizing.

## Basic principles of sound synthesis in soundgen
Soundgen's credo is to start with a few control parameters (e.g., the intonation contour, the amount of noise, the number of syllables and their duration, etc.) and to generate a corresponding audio stream, which will sound like a biological vocalization (a moan, a laugh, etc). The core algorithm for generating a single voiced segment implements the standard source-filter model, which is the cornerstone of phonetics (Fant 1971). The voiced component is generated as a sum of sine waves and the noise component as filtered white noise, and both components are then passed through a frequency filter simulating the effect of human vocal tract. This process can be conceptually divided into three stages

1.  Generation of the harmonic component (glottal source).
    At this crucial stage, we "paint" the spectrogram of the glottal source based on the desired intonation contour and spectral envelope by specifying the frequencies, phases, and amplitudes of a number of sine waves, one for each harmonic of the fundamental frequency. If needed, we also add stochastic and non-linear effects at this stage: jitter and shimmer (random fluctuation in frequency and amplitude), subharmonics, slower random drift of control parameters, etc. Once the spectrogram "painting" is complete, we synthesize the corresponding waveform by generating and adding up as many sine waves as there are harmonics in the spectrum. 
    
    Note that soundgen currently implements only sine wave synthesis. This is different from modeling glottal cycles themselves, as in phonetic models and some popular text-to-speech engines (e.g. Klatt 1980). In future versions of soundgen there may be an option to use a particular parametric model of the glottal cycle as excitation source as an alternative to generating a separate sine wave for each harmonic.

1.  Generation of the noise component (aspiration, etc.).
    In addition to harmonic oscillations of the vocal cords, there are other sources of excitation, which may be generated as some form of noise. For example, aspiration noise may be synthesized as white noise with rolloff -6 dB/octave (Klatt 1990) and added to the glottal source before formant filtering. It is similarly straightforward to add other types of noise, which may originate higher up in the vocal tract and thus display a different formant structure from the glottal source (e.g., high-frequency hissing, broadband clicks for tongue smacking, etc.) 
    
    Some form of noise is synthesized in most sound generators. In soundgen noise is created in the frequency domain (i.e., as a spectrogram) and then converted into a time sequence via inverse FFT.
    
1.  Spectral filtering (formants and lip radiation).
    The vocal tract acts as a resonator that modifies the source spectrum by amplifying certain frequencies and dampening others. In speech, time-varying resonance frequencies (formants) are responsible for the distinctions between different vowels, but formants are also ubiquitous in animal vocalizations. Just as we "painted" a spectrogram for the acoustic source in (1), we now "paint" a spectral filter with a specified number of stationary or moving formants. We then take a fast Fourier transform (FFT) of the generated waveform to convert it back to a spectrogram, multiply the latter by our filter, and then take an inverse FFT to go back to the time domain. This filtering can be applied to harmonic and noise components separately or - for noise sources close to the glottis - the harmonic component and the noise component can be added first and then filtered together. 
    
    Note that this FFT-mediated method of adding formants is different from the more traditional convolution, but with multiple formants it is both considerably faster and (arguably) more intuitive.

Having briefly looked at the fundamental principles of sound generation, we proceed to control parameters. The aim of the following presentation is to offer practical tips on using soundgen. For further information on more fundamental principles of acoustics and sound synthesis, you may find the vignettes in `seewave` very helpful, and look out for the upcoming book on sound synthesis in R by Jerome Sueur, the author of the `seewave` package.



# Using soundgen
## Where to start
To generate a sound, you can either type `soundgen_app()` to open an interactive Shiny app or call `soundgen()` with manually specified parameters. 

The basic workflow in the Shiny app is as follows:

1.  Set parameters in the tabs on the left (see the sections below for details). You can also start with a preset that resembles the sound you want and then fine-tune control parameters.
1.  Check the preview plots, tables of anchors etc to ensure you get what you want.
1.  Click Generate to create a WAV file, play it (NB: no sound in Opera browser!), and display a spectrogram of the generated sound.
1.  Save the WAV file or go back to (1) to make further adjustments. 

*TIP The interactive app `soundgen_app()` gives you the exact R code for calling `soundgen()`, which you can copy-paste into your R environment and generate manually the same sound as the one you have created in the app. If you find yourself in doubt about the right format for specifying a parameter for `soundgen()`, you can use the app first, then copy-paste the code into your R console, and modify it as needed.*

## Syllables
If you need to generate a single syllable without pauses, the only parameter you have to set is `sylLen` ("Syllable length, ms" in the app). For a bout of several syllables, you have two options:

1.  Set `nSyl` ("Number of syllables" in the app). Breathing is then allowed to fill in the pauses (if breathing is longer than the voiced part), and you can specify an amplitude contour, intonation contour, and formant transitions that will span the entire bout. For ex., if the vowel sequence in a three-syllable bout is “uai”, the output will be approximately “[u] – pause – [a] – pause – [i]”. 
```{r}
library(soundgen)
s = soundgen(exactFormants = 'uai', nSyl = 3, play = TRUE)
# "play = TRUE": in the rest of the vignette sound output is muted ("play = FALSE"). 
# To play/repeat, type "playme(s)" instead
```
1.  Set `repeatBout` ("Repeat bout # times" in the app). This is the same as calling `soundgen` repeatedly with the same settings or clicking the Generate button in the app several times. If temperature = 0, you will get exactly the same sound repeated each time, otherwise some variation will be introduced. For the same “uai” example, the output will be “[uai] – pause – [uai] – pause – [uai]”.
```{r}
s = soundgen(exactFormants = 'uai', repeatBout = 3)
# playme(s)
```

## Intonation
When we hear a tonal sound such as someone singing, one of its most salient characteristic is intonation or, more technically, the contour of fundamental frequency (F0), or, even more technically, the contour of the spectral band which is perceived to correspond to the fundamental frequency (pitch). Soundgen literally generates a sine wave corresponding to F0 and several higher harmonics, so F0 is straightforward to implement. However, how can its contour be specified with as few parameters as possible? The solution adopted in soundgen is to take one or more anchors as input and generate a smooth contour that passes through all anchors. 

The function that generates these smooth contours is `getSmoothContour()`. You do not have to call it explicitly, but sometimes it can be helpful to do so in order to visualize the curve implied by your anchors. Anchors are points defined by their time (ms) and value (in the case of pitch, this is frequency in Hz). Time anchors can range 0 to 1, or they can be specified in ms – it makes no difference since the sound is rescaled to match the duration `sylLen`.

For example, say we want F0 to increase linearly from 350 to 700 Hz. Time anchors can then be specified as `c(0, 1)`, and the arguments `len` and `samplingRate` together determine the duration in s: `len / samplingRate = time_s`. Values are processed on a logarithmic (musical) scale if `thisIsPitch` is `TRUE`: observe that C4 (note C of octave 4), C5, and C6 are equidistant on the right-hand Y axis.

```{r fig.width = 4, fig.height = 4}
a = getSmoothContour(anchors = data.frame(time = c(0, 1), value = c(350, 700)),
  len = 7000, thisIsPitch = TRUE, plot = TRUE, samplingRate = 3500)
```

A sound with this intonation can be generated as follows (note that we have to specify what aspect of the sound the anchors apply to - `pitchAnchors`):
```{r}
sound = soundgen(sylLen = 2000, 
                 pitchAnchors =  data.frame(time = c(0, 1), value = c(350, 700)))
```
To get an inverted U shape with ascending-descending intonation instead of a linearly increasing F0, we add a third anchor:
```{r fig.width = 4, fig.height = 4}
a = getSmoothContour(anchors = data.frame(time = c(0, .5, 1), value = c(350, 700, 350)),
  len = 7000, thisIsPitch = TRUE, plot = TRUE, samplingRate = 3500)
```

To draw F0 contour in the app, use "Main / Intonation syllable" tab and click the intonation plot to add anchors. Soundgen then generates a smooth curve through these anchors. If you click the plot close to an existing anchor, it moves to where you clicked; if you click far from any existing anchor, a new anchor is added. To remove an anchor, double-click close to it. To go back to a straight line, click the button labeled “Flatten pitch contour”.

*TIP The same principles apply to all contour plots in the app (amplitude, mouth opening, and noise contours). Note also that all contours are rescaled when the duration changes, with the single exception of negative time anchors for noise (i.e. the length of pre-syllable aspiration does not depend on syllable duration).*

*TIP Given the same anchors, the shape of the resulting curve depends on syllable duration. That's because the amount of smoothing is adjusted automatically if you change syllable duration. Double-check that all your contours still look reasonable if you change the duration!*

If the bout consists of several syllables (`nSyl > 1`), you can also specify the overall intonation over several syllables using `pitchAnchors_global` (app: "Main / Intonation global"). The global intonation contour specifies the deviation of pitch per syllable from the main pitch contour in semitones, i.e. 12 semitones = 1 octave. In other words, it shows how much higher or lower the average pitch of each syllable is compared to the rest of the syllables. For ex., we can generate five seagull-like sounds, which have the same intonation contour within each syllable, but which vary in average pitch spanning about an octave in an inverted U-shaped curve:
```{r fig.width = 5, fig.height = 5}
s = soundgen(nSyl = 5, sylLen = 200, pauseLen = 140, plot = TRUE,
                 pitchAnchors = data.frame(time = c(0, 0.65, 1), value = c(977, 1540, 826)),
                 pitchAnchors_global = data.frame(time = c(0, .5, 1), value = c(-6, 7, 0)))
# playme(s)
```

*TIP Calling `soudngen` with argument `plot = TRUE` produces a spectrogram using a function from soundgen package, `spec`. Type `?spec` and see the vignette on acoustic analysis for plotting tips and advanced options. Or you can plot the waveform produced by `soundgen` using another function of your choice, e.g. `seewave::spectro`*

## Hyper-parameters
### Stochasticity = “temperature”
It is a basic principle of soundgen that random variation can be introduced in the generated sound. This behavior is controlled by a single high-level parameter, `temperature`. If `temperature = 0`, you will get exactly the same sound by executing the same call to `soundgen` repeatedly. If `temperature > 0`, each generated sound will be somewhat different, even if all the control parameters are exactly the same. In particular, positive temperature introduces fluctuations in syllable structure, all contours (intonation, breathing, amplitude, mouth opening), and many effects (jitter, subharmonics, etc). It also "wiggles" user-specified formants and adds new formants above the specified ones at a distance calculated based on the vocal tract length.

Code example : 
```{r, eval = FALSE}
# the sound is a bit different each time, because temperature is above zero
for (i in 1:3) {
  soundgen(temperature = 0.3, play = TRUE)
}
```

If you don't want stochastic behavior, set temperature to zero. But note that some effects, notably jitter and subharmonics, will then be added in an all-or-nothing manner: either to the entire sound or not at all. You can also change the extent to which temperature affects different parameters (e.g., if you want more variation in intonation and less variation in syllable structure). To do so, use `tempEffects`, which is a list of scaling coefficients that determine how much different parameters vary at a given temperature. You can check their names and default settings by looking at the documentation of soundgen: type `?soundgen`.
```{r}
# despite the high temperature, formants are not going to vary at all, 
# while temporatal structure will be more variable than the default
for (i in 1:3) {
  s = soundgen(nSyl = 3, temperature = .3, tempEffects = list(sylLenDep = .5, formDrift = 0))
  # playme(s)
}
```

### Other hypers
To simplify usage, there are a few other hyper-parameters that are not strictly necessary to produce the full range of sounds, but provide convenient shortcuts by making it possible to control several low-level parameters at once in a coordinated manner. Hyper-parameters are marked "hyper" in the Shiny app. 

For example, to imitate the effect of varying body size, you can use `maleFemale`:
```{r, eval = FALSE}
mf = c(-1,  # male: 100% lower F0, 25% lower formants, 25% longer vocal tract
       0,   # neutral (default)
       1)   # female: 100% higher F0, 25% higher formants, 25% shorter vocal tract
for (i in mf){
  s = soundgen(maleFemale = i, exactFormants = NA, vocalTract = 25, play = TRUE)
  # Since `exactFormants` are not specified, but temperature is above zero, a 
  # schwa-like sound with approximately equidistant formants is generated using
  # `vocalTract_length` (cm) to calculate the expected formant dispersion.
}
```

To change the basic voice quality along the breathy-creaky continuum, use `creakyBreathy`. It affects the rolloff of harmonics, the type and strength of pitch effects (jitter, subharmonics), and the amount of aspiration noise. For example:
```{r, eval = FALSE}
cb = c(-1,  # max creaky
       -.5, # moderately creaky
       0,   # neutral (default)
       .5,  # moderately breathy
       1)   # max breathy
for (i in cb){
  par(mfrow = c(1, 3))
  soundgen(creakyBreathy = i, play = TRUE)
  par(mfrow = c(1, 1))
}
```

## Amplitude contours
Use `amplAnchors` and `amplAnchors_global` to modulate the amplitude (loudness) of an individual syllable or a polysyllabic bout, respectively. In the app, they are found under "Main / Amplitude syllable"" and "Main / Amplitude global". Note that they both affect only the voiced component. In contrast, `attackLen` ("Attack length, ms" in the app) and `trill_dep` ("Trill") affect both the voiced and the unvoiced components. 
```{r fig.width = 5, fig.height = 5}
# each syllable has a 20-dB dip in the middle, and there is an overall fade-out over the entire bout
s = soundgen(nSyl = 4, plot = TRUE, osc = TRUE,
             amplAnchors = data.frame(time = c(0, .5, 1), value = c(120, 100, 120)),
             amplAnchors_global = data.frame(time = c(0, 1), value = c(120, 0)))
# playme(s)
```
Rapid amplitude modulation imitating the uvular trill (as in French [r]) is implemented by multiplying the synthesized waveform by a sine wave:
```{r fig.width = 5, fig.height = 5}
s = soundgen(sylLen = 1000, exactFormants = NA,
             trillDep = .5,   # halves the amplitude at troughs
             trillFreq = 35,  # amplitude modulation with frequency 35 Hz
             plot = TRUE, osc = TRUE)
# playme(s)
```

## Vowel quality (formants)
### Vowel presets
Argument `exactFormants` (tab "Voiced / Vowel" in the app) sets the formants – frequency bands used to filter the excitation source. If this sounds fancy, think of the sort of equalizer you might use to boost the bass when listening to a song. Some frequencies are amplified, others are dampened. When appropriate filters are applied to a tonal sound like human voice, we perceive different vowels.

For human voices (callers M1 and F1 in the app), you can specify a string of vowels from the dictionary of presets. When you call `soundgen` with `exactFormants = 'ai'` or some other character string, the values are taken from `presets$M1$Formants` and `presets$F1$Formants`. Formants can remain the same throughout the vocalizations, or they can move. For example, `exactFormants = 'ai'` produces a sound that goes smoothly from [a] to [i], while `exactFormants = 'aaai'` produces mostly [a] with a quick transition to [i] at the very end. Argument `formantStrength` ("Formant prominence" in the app) adjusts the overall effect of all formant filters at once.

### Manual formants
Presets give you some rudimentary control over vowels. More subtle control is necessary for animal sounds, as well as for human vowels that are not included in the presets dictionary. For such cases you will have to specify the actual frequency, amplitude, and bandwidth of each formant manually, as well as time stamps for each value. If you want stationary formants, type `time = 0` for each formant. For moving formants, you can specify values at different time points, where time varies from 0 to 1 (to be scaled appropriately depending on the length of sound). For example, the following example uses two moving formants with frequency, amplitude, and bandwidth specified at the beginning (time 0) and end (time 1) of the syllable:
```{r fig.width = 4, fig.height = 4}
exactFormants = list(
  f1 = data.frame(time = c(0, 1), freq = c(300, 900), amp = c(30, 10), width = 120),
  f2 = data.frame(time = c(0, 1), freq = c(2500, 1500), amp = 30, width = c(0, 240)))

# normally you would simply feed this list into soundgen(), but sometimes
# it may be helpful to plot the spectral filter implied by your formants:
s = getSpectralEnvelope(nr = 1024,  # freq bins in FFT frame (window_length / 2)
                        nc = 50,    # time 
                        samplingRate = 16000, 
                        exactFormants = exactFormants,
                        plot = TRUE, 
                        dur_ms = 1500,   # just an example
                        colorTheme = 'seewave',
                        rolloff_lipRad = 6) 
# Note that lip radiation is specified in exactFormants together with formants! 
# This has the effect of amplifying higher frequencies to mimic lip radiation

# synthesize a sound using this spectral filter:
s = soundgen(exactFormants = exactFormants)
# playme(s)
```

*TIP When using the app, you can start with a preset by typing in a vowel string, and then you can modify it. This way you don't have to remember the right format. If you edit the list of formants and nothing in the sound seems to be changing, there may be a misprint, missing comma, etc. To make sure your formants are correctly specified, you can also plot them directly in R by calling `getSpectralEnvelope` (see above)*

For even more advanced spectral filters, you can specify both formants and antiformants. This may be useful if you want to create a nasalized sound. The numbering of formants is arbitrary, as long as they are arranged in the right order: if you need to insert a new formant between F1 and F2, call it "f1.5" or something like that. For example, a slow transition from [a] to [a nasalized] might be coded as follows (note that formant f1.7 has negative amplitude, so that f1.5 and f1.7 form a pole-zero pair):
```{r fig.width = 5, fig.height = 5}
s = soundgen(sylLen = 1500,
             pitchAnchors = data.frame(time = c(0, 1), value = c(140, 140)), 
             exactFormants = list(
               f1   = data.frame(time = c(0, 1), freq = c(880, 900), 
                                 amp = c(40,20), width = c(80,120)), 
               f1.5 = data.frame(time = c(0, 1), freq = 600, 
                                 amp = c(0, 30), width= 80), 
               f1.7 = data.frame(time=c(0, 1), freq = 750, 
                                 amp = c(0,-80), width = 80), 
               f2   = data.frame(time = c(0, 1), freq = c(1480, 1250), 
                                 amp = c(40, 20), width = c(120, 200)), 
               f3   = data.frame(time=c(0, 1), freq = c(2900, 3100), 
                                 amp = 40, width = 200)))
# The spectrogram of the resulting sound clearly shows the zero-pole pair:
spec(s, samplingRate = 16000, ylim = c(0, 4), contrast = .5, 
     windowLength = 10, step = 5, colorTheme = 'seewave')
# playme(s)
```

### Mouth opening
A convenient shortcut for manipulating formants without coding all transitions by hand is provided by `mouthAnchors` argument (in the app, tab "Main / Mouth opening"). This can be thought of as a hyper-parameter offering an easy way to define moving formants within a syllable: formants go down as the mouth closes and rise as it opens. In addition, lip radiation is removed when the mouth is completely closed, and the vowel is automatically nasalized. This feature can save you a lot of manual coding, but it is basically redundant – you can always achieve the same effect by specifying moving formants manually in `exactFormants`. Here is a simple example, with the mouth gradually opening and closing again:
```{r fig.width = 5, fig.height = 5}
s = soundgen(sylLen = 2500,
             pitchAnchors = data.frame(time = c(0, 1), value = c(140, 140)), 
             mouthAnchors = list(time = c(0, .3, .75, 1), value = c(0, 0, .7, 0)))
# playme(s)
spec(s, samplingRate = 16000, ylim = c(0, 4), contrast = .5, windowLength = 10, step = 5,
     colorTheme = 'seewave', osc = TRUE)
```

## Source spectrum
Soundgen produces tonal sounds by means of generating a separate sine wave for each harmonic. However, it is very tricky to choose the appropriate strength of each harmonic. The simplest solution is to make each higher harmonic slightly weaker than the previous one, say by setting a fixed exponential decay rate from lower to higher harmonics. The corresponding parameter in soundgen is `rolloff` (in the app, "Source rolloff, dB/octave"). Unfortunately, this is often not really good enough, necessitating several more control parameters. 

Soundgen allows a lot of flexibility when specifying source spectrum. You can change the basic rolloff of harmonics per octave, adjust rolloff depending on F0, add parabolic terms that affect the first few harmonics, etc. Working from R console, the relevant function is `getRolloff`. Its arguments are well-documented: type `?getRolloff` for help. Here is just a single example:
```{r}
# strong F0, shoulder around 2 kHz
r = getRolloff(rolloff = -20, rolloffAdjust_per_octave = -3,
               rolloffAdjust_quadratic = -10, rolloffAdjust_quadratic_nHarm = 13, 
               pitch_per_gc = 170, plot = TRUE)

# to generate the corresponding sound:
s = soundgen(rolloff = -20, rolloffAdjust_per_octave = -3,
             olloffAdjust_quadratic = -10, rolloffAdjust_quadratic_nHarm = 13,
             pitchAnchors = data.frame(time = 0, value = 170))
# playme(s)
```

In the app the relevant parameters are found in the tab "Voiced / Source spectrum". These settings may at first seem obscure and are hard to optimize, especially since their effect also depends on formants, but they have a major impact on timbre (sound quality). To develop an intuition for source spectrum settings, I recommend practicing with disabled formants in the app (set "Formants prominence" under "Main / Vowel" to 0). This way you can isolate the effects of source spectrum and use the preview plot for instant feedback – it shows the rolloff for the lowest and the highest pitch in your intonation contour.

## Pitch effects
Several modifications of the voiced component can basically make the sound appear harsh. In biological sounds this happens on the level of individual glottal cycles. The corresponding effects in soundgen are loosely grouped under the label "pitch effects". Soundgen can add subharmonics (vocal fry), jitter (random fluctuation in fundamental frequency per glottal cycle), and shimmer (random fluctuation in amplitude per glottal cycle). 

Technically, no actual glottal cycles are involved, since soundgen creates continuous sine waves, but it divides F0 contours into chunks approximating glottal cycles. Jitter and shimmer are then created by adding random noise to the periods and amplitudes, respectively, of these "fake" glottal cycles. Subharmonics could be created by adding rapid frequency modulation to F0 contour, but soundgen uses a different - slightly hacky, but extremely powerful - technique of literally setting up an additional sine wave for each subharmonic based on the desired frequency of subharmonics (`subFreq`). The amplitude of each subharmonic is a function of its distance from the nearest harmonic from F0 stack and the desired width of sidebands (`subDep`). This way we can create sidebands that vary naturally as F0 changes over time, producing bifurcations, and dynamically vary the nature of subharmonic regimes. The main limitation of this approach is that it is too computationally costly to generate variable numbers of subharmonics for the entire bout. The solution currently adopted in soundgen is to break longer sounds into so-called "epochs" with a constant number of subharmonics in each. The epochs are synthesized separately, trimmed to the nearest zero crossing, and then glued together with a rapid cross-fade. This is suboptimal, since it shortens the sound and may introduce audible artifacts at transitions between epochs. `shortestEpoch` controls the approximate minimum length of each epoch. Longer epochs minimize problems with transitions, but the behavior of subharmonics then becomes less variable, since their number is constrained to be stable within each epoch.

To add pitch effects, you can use just two parameters – `pitchEffects_amount` and `pitchEffects_intensity`  – to regulate approximately how harsh the sound becomes and what proportion of the sound is affected, respectively. However, for best results it is advisable to set advanced settings manually (see below). At temperature > 0, `pitchEffects_amount` creates a random contour that divides each syllable into epochs defined by their noise regime:

1.  No pitch effects. 
    If pitchEffects_amount = 0%, the whole syllable is in pitch regime 1.
1.  Subharmonics only. 
    Note that subharmonics are only added to segments with subFreq < F0 / 2.
1.  Subharmonics and jitter. 
    If pitchEffects_amount = 100%, the whole syllable is in pitch regime 3. 

`pitchEffects_intensity` is a hyper-parameter that adjusts several settings at once, making the voice harsher in pitch regimes 2 and 3, but without affecting the balance between regimes.

Moving on to advanced pitch effects settings, `subFreq` ("Subharmonic frequency, Hz" in the app) and `subDep` ("Width of sidebands, Hz") define the properties of subharmonics in pitch regimes 2 and 3. Say your vocalization has a relatively flat intonation contour with a fundamental frequency of about 800 Hz, and you want to add a single subharmonic (G0). You then set the expected subharmonic frequency to 400 Hz. Since G0 is forced to be an integer fraction of F0 at all time points, it will not be exactly 400 Hz, but it will produce a single subharmonic at F0 / 2 (as long as F0 stays close to 800 Hz: if it goes up to 1200 Hz, you will get two subharmonics instead, since 1200 / 400 = 3). The width of sidebands defines how quickly the energy of subharmonics decays at a remove from the nearest harmonic. For example, our single subharmonic is audible but weak at sideband width = 150 Hz, while it becomes strong enough to be perceived as a new fundamental at sideband width = 400 Hz, effectively halving the pitch:
```{r fig.width = 5, fig.height = 5}
s1 = soundgen(subFreq = 400, subDep = 150, pitchEffects_amount = 100,
              jitterDep = 0, shimmerDep = 0, temperature = 0, 
              sylLen = 500, pitchAnchors = data.frame(time=c(0, 1), value = c(800, 900)),
              plot = TRUE)
# playme(s1)
s2 = soundgen(subFreq = 400, subDep = 400, pitchEffects_amount = 100,
              jitterDep = 0, shimmerDep = 0, temperature = 0, 
              sylLen = 500, pitchAnchors = data.frame(time=c(0, 1), value = c(800, 900)),
              plot = TRUE)
# playme(s2)
```

Sidebands may be easier to understand for high-pitched sounds with low subharmonic frequencies. For example, chimpanzees emit piercing screams with narrow subharmonic bands. If we set `subFreq` to 75 Hz and `subDep` to 130 Hz, subharmonics literally form a band around each harmonic of the main stack, creating very distinct, immediately recognizable sound quality:
```{r fig.width = 5, fig.height = 5}
s1 = soundgen(subFreq = 75, subDep = 130, pitchEffects_amount = 100,
              jitterDep = 0, shimmerDep = 0, temperature = 0, 
              sylLen = 800, plot = TRUE,
              pitchAnchors = data.frame(time=c(0, .3, .9, 1), 
                                        value = c(1200, 1547, 1487, 1154)))
# playme(s1)
```

As for jitter in pitch regime 3, it wiggles both F0 and G0 harmonic stacks, blurring the spectrum. Parameter `jitterDep` ("Jitter depth, semitones" in the app) defines how much the pitch fluctuates, while`jitterLen` ("Jitter period, ms") defines how rapid these fluctuations are. Slow jitter with a period of ~50 ms creates the effect of a shaky, unsteady voice. It may sound similar to a vibrato, but jitter is irregular. Rapid jitter with a period of ~1 ms may be used to imitate deterministic chaos, which is found in voiced but highly irregular animal sounds such as barks, roars, noisy screams, etc. 
```{r}
# To get jitter without subharmonics, set `temperature = 0, subDep = 0` 
# and specify the required jitter depth and period
s1 = soundgen(jitterLen = 50, jitterDep = 1,  # shaky voice
              sylLen = 1000, subDep = 0, pitchEffects_amount = 100,
              pitchAnchors = data.frame(time = c(0, 1), value = c(150, 170)))
# playme(s1)
s2 = soundgen(jitterLen = 1, jitterDep = 1,  # harsh voice
              sylLen = 1000, subDep = 0, pitchEffects_amount = 100,
              pitchAnchors = data.frame(time = c(0, 1), value = c(150, 170)))
# playme(s2)
```

To get both jitter and subharmonics, set both manually and/or use pitchEffects_amount close to 100% with temperature > 0. For example, barks of a small, annoying dog can be roughly approximated with this minimal code:
```{r}
s = soundgen(repeatBout = 2, sylLen = 140, pauseLen = 100, vocalTract = 8,
             pitchAnchors = list(time = c(0, 0.52, 1), value = c(559, 785, 557)), 
             pitchEffects_amount = 100, jitterDep = 1, subDep = 60, 
             mouthAnchors = list(time = c(0, 0.5, 1), value = c(0, 0.5, 0)))
# playme(s)
```

### Vibrato
Technically also a pitch effect, vibrato has nothing to do with subharmonics and jitter. It simply adds a *regular* (in contrast to irregular jitter and temperature-related random drift), perfectly sinusoidal frequency modulation to F0 contour:
```{r}
# 5-Hz vibrato 1 semitone in depth
s1 = soundgen(vibratoDep = 1, vibratoFreq = 5, sylLen = 1000, 
              pitchAnchors = data.frame(time = c(0, 1), value = c(300, 280)))
# playme(s1)
# slower (3 Hz) and deeper (3 semitones) vibrato
s2 = soundgen(vibratoDep = 3, vibratoFreq = 3, sylLen = 1000, 
              pitchAnchors = data.frame(time = c(0, 1), value = c(300, 280)))
# playme(s2)
```

## Unvoiced component = noise
In addition to the tonal (harmonic, voiced) component, which is synthesized as a stack of harmonics (sine waves), soundgen produces broad-spectrum noise (unvoiced component). This noise can be added to the voiced component to create breathing, snuffing, hissing, etc. This can be done in two ways:

1.  Breathing. This noise type is generated as white noise with spectral rolloff given by `rolloff_noise` ("Noise rolloff, dB/octave" in the app). It is added to the voiced component *before* formant filtering. As a result, it follows exactly the same formant structure as the voiced component, and you cannot modify its spectrum beyond the basic rolloff setting. This is useful for adding noise that originates deep in the throat, close to the vocal cords. To generate breathing, specify `noiseAnchors`, but leave `exactFormants_noise` blank (NA, which is its default value). Soundgen then assumes that noise should have the same formant structure as the voiced part.

1.  Any other noise type is added to the voiced component *after* formant filtering, and therefore this noise can be filtered independently of the voiced component. To generate such noise, you can use one of the available presets in the app (for now, only a few human consonants) or specify the formants for the unvoiced component manually in exactly the same format (`exactFormants_noise`) as for the voiced component (`exactFormants`). 

In the shiny app, the tab "Unvoiced / Noise timing" is for specifying the amplitude contour of the unvoiced component. It shows the timing of noise relative to the voiced component of a typical syllable. Note that noise is allowed to fill the pauses between syllables, but not between bouts. For example, in this two-syllable bout noise carries over after the end of each voiced component, since syllable duration is 120 ms and the last breathing time anchor is 209 ms:
```{r fig.width = 5, fig.height = 5}
s = soundgen(nSyl = 2, sylLen = 120, pauseLen = 120, plot = TRUE, osc = TRUE,
             temperature = 0, rolloff_noise = -5, 
             noiseAnchors = data.frame(time = c(39, 56, 209), value = c(-120, -10, -120)),
             exactFormants = list(f1 = data.frame(time = c(0, 1), freq = c(860, 530), 
                                                  amp = 30, width = c(120, 50)), 
                                  f2 = data.frame(time = c(0, 1), freq = c(1280, 2400), 
                                                  amp = 40, width = c(120, 300))),
             exactFormants_noise = list(f1 = data.frame(time = 0, freq = 420, 
                                                        amp = 20, width = 150),
                                        f2 = data.frame(time = 0, freq = 1200, 
                                                        amp = 50, width = 250)))
# playme(s)
```
Note that both timing and amplitude of noise anchors are defined relative to the voiced component. Time anchors for noise MUST be specified in ms (unlike all the other contours, which accept time anchors on any arbitrary scale, say 0 to 1). If the noise starts before the voiced part, the first time anchor will be negative. The app will give you an easy preview. From R console, you can also preview the noise amplitude contour implied by your anchors by calling `getSmoothContour`:
```{r fig.width = 4, fig.height = 4}
a = getSmoothContour(anchors = data.frame(time = c(-50, 200, 300), 
                                          value = c(-120, 20, -120)),
                     voiced = 200, plot = TRUE, ylim = c(-120, 40), main = '')
```

*TIP: if the voiced part is shorter than `permittedValues['sylLen', 'low']`, it is not synthesized at all, so you only get the unvoiced component (if any)*

Note that at temperature > 0 breathing noise is enriched with stochastically added formants, just like the voiced component. To create simple sighs, you can just specify the length of your creature's vocal tract:
```{r}
s1 = soundgen(vocalTract = 17.5,  # human throat
              sylLen = 10, exactFormants = NULL, attackLen = 200,
              noiseAnchors = list(time = c(-8, 813), value = c(40, 40)))
# playme(s1)
s2 = soundgen(vocalTract = 30,    # a large animal
              sylLen = 10, exactFormants = NULL, attackLen = 200,
              noiseAnchors = list(time = c(-8, 813), value = c(40, 40)))
# playme(s2)
```

# Morphing two sounds
Sometimes it is desirable to combine characteristics of two different stiimuli, producing some kind of intermediate form - a hybrid or blend of both originals. This technique is called morphing, and it is often and successfully used with visual stimuli, but not so often with sounds, because it turns out to be rather tricky to morph audio. Since soundgen creates sounds parametrically, however, morphing becomes much more straightforward: all we need to do is define the rules for interpolating between all control parameters. For example, say we have sound A (100 ms) and sound B (500 ms), which only differ in their duration. To morph them, we could generate five otherwise identical sounds that are 100, 200, 300, 400, and 500 ms long, giving us the originals and three equidistant intermediate forms - that is, if we assume that linear interpolation is the natural way to make sounds of different duration equidistant.

In practice this assumption is often unwarranted. For example, the natural scale for pitch is log-transformed: the perceived distance between 100 Hz and 200 Hz is 12 semitones, while from 200 Hz to 300 Hz it is only 7 semitones. To make pitch values equidistant, we would need to think in terms of semitones, not Hz. For other soundgen parameters it is hard to make an educated guess about the natural scale, so the most appropriate interpolation rules remains obscure. For best results, morphing should be performed by hand, pre-testing each parameter of interest and creating the appropriate formulas for each morph. However, for a "quick fix" there is an in-built function, `morph`.

`morph` takes two calls to `soundgen` (as a character string or list of arguments) and creates several morphs using linear interpolation for all parameters except pitch and formant frequencies, which are log-transformed prior to interpolation and then exponentiated to go back to Hz. The morphing algorithm can also deal with arbitrary contours, either by taking a weighted mean of each curve (`method = 'smooth'`) or by attempting to match and morph individual anchors (`method = 'perAnchor'`):

```{r fig.width = 7.5, fig.height = 3}
a = data.frame(time=c(0, .2, .9, 1), value=c(100, 110, 180, 110))
b = data.frame(time=c(0, .3, .5, .8, 1), value=c(300, 220, 190, 400, 350))
par(mfrow = c(1, 3))
plot (a, type = 'b', ylim = c(0, 500), main = 'Original curves')
points (b, type = 'b', col = 'blue')
m = soundgen:::morphDF(a, b, nMorphs = 15, method = 'smooth', 
                       plot = TRUE, main = 'Morphing curves')
m = soundgen:::morphDF(a, b, nMorphs = 15, method = 'perAnchor', 
                       plot = TRUE, main = 'Morphing anchors')
par(mfrow = c(1, 1))
```

Here is an example of morphing the default neutral [a] into a dog barking:
```{r}
m = morph(formula1 = list(repeatBout = 2),
          # equivalently: formula1 = 'soundgen(repeatBout = 2)',
          formula2 = presets$Misc$Dog_bark,
          nMorphs = 5, playMorphs = TRUE)
# use $formulas to access formulas for each morph, $sounds for waveforms
# m$formulas[[4]]
# playme(m$sounds[[3]])
```


